Об оценках на диаметр плоских множеств с целочисленными расстояниями полуобщего положения \footnote{Исследование выполнено за счет гранта Российского научного фонда (проект 19-11-00197).}

### 2019-11-02 semi-general-suplinear-bound/CHEBSB_Avdeev.tex
Author thanks  Dr. Usachev for remarks on English

\Abstract
{%Аннотация статьи на русском языке 150-250 слов с учетом ключевых слов
Множество точек $M$ на плоскости называется плоским множеством с целочисленными расстояниями,
если все расстояния между точками $M$ суть целые числа,
и при этом $M$ не содержится ни в какой прямой.
Говорят, что плоское множество с целочисленными расстояниями есть множество полуобщего положения,
если никакие три его точки не лежат на одной прямой.
Известная оценка снизу на диаметр плоского множества с целочисленными расстояниями
линейна относительно его мощности.
Ранее не были известны отдельные оценки снизу на диаметр плоских множеств с целочисленными расстояниями полуобщего положения
заданной мощности
(известная конструктивная оценка сверху на диаметр плоских множеств с целочисленными расстояниями
использует как раз множества полуобщего положения).
В статье доказывается надлинейная оценка снизу
на диаметр плоского множества с целочисленными расстояниями полуобщего положения
(полиномиальная с показателем $5/4$).
Доказательство основано на относительно большом количестве лемм и наблюдений,
включая результаты Солимоси
из статьи, в которой была впервые доказана линейная оценка снизу
на диаметр плоских множеств с целочисленными расстояниями.
Полученная в данной статье оценка всё ещё не точна.
}
{%Аннотация статьи на английском языке
A point set $M$ in the Euclidean plane is called a planar integral point set if all the distances between the
elements of $M$ are integers, and $M$ is not situated on a straight line.
A planar integral point set is called to be in semi-general position, if it does not contain collinear triples.
The existing lower bound for mininum diameter of a planar integral point set is linear with respect to its cardinality.
There were no known special diameter bounds for planar integral point sets in semi-general position
(the known upper bound for planar integral point sets is constructive
and provides planar integral point sets in semi-general position).
We prove a new lower bound for mininum diameter of planar integral point sets in semi-general position
that is better than linear (polynomial of power $5/4$).
The proof is based on several lemmas and observations, including the ones provided by Solymosi
while proving the first linear lower bound for diameter of a planar integral point set.
We have to admit that our bound is still not tight.
}

@@

Since every integral point set can obviously be dilated to a set of larger diameter,
minimal possible diameters of sets of given cardinality are in the focus.
To be precise,
the following function was introduced~\cite{kurz2008bounds,kurz2008minimum}:
\begin{equation}
	d(2,n) = \min_{M\in\mathfrak{M}(2,n)} \operatorname{diam} M
	.
\end{equation}

It turned out to be very easy to construct a planar integral point set of $n$ points with $n-1$ collinear ones and one point out of the line
(so-called \textit{facher} sets);
the same holds for 2 points out of the line (we refer the reader to~\cite{antonov2008maximal}, where some of such sets are called \textit{crabs})
and even for 4 points out of the line~\cite{huff1948diophantine}.
For $9\leq n\leq 122$, the minimal possible diameter is achieved at a facher set~\cite{kurz2008bounds}.

@@

Furthermore, the constructions of integral point sets in semi-general position of arbitrary cardinality
appeared~\cite{harborth1993upper};
such sets are situated on a circle.
Also, there is a sophisticated construction of a circular integral point set of arbitrary cardinality
that gives the possible numbers of odd integral distances
between points in the plane~\cite{piepmeyer1996maximum}.

\begin{Definition}
	A set $M\in\overline{\mathfrak{M}}(2,n)$ is called to be in \textit{general position},
	if no four points of $M$ are concyclic.
	The set of all planar integral point sets in general position
	is denoted by $\dot{\mathfrak{M}}(2,n)$.
\end{Definition}

It remains unknown if there are integral points sets in general position of arbitrary cardinality;
however, some sets $M\in \dot{\mathfrak{M}}(2,7)$ are known~\cite{kreisel2008heptagon,kurz2013constructing}.

The inequality
\begin{equation*}
	d(2,n) \leq \overline{d}(2,n) \leq \dot{d}(2,n)
	,
\end{equation*}
where
$
	\overline{d}(2,n) = \min_{M\in\overline{\mathfrak{M}}(2,n)} \operatorname{diam} M
$
and
$
	\dot{d}(2,n) = \min_{M\in\dot{\mathfrak{M}}(2,n)} \operatorname{diam} M
$,
is obvious; however, a more interesting relation holds:
\begin{equation*}
	c_1 n \leq d(2,n) \leq \overline{d}(2,n) \leq n^{c_2 \log \log n}
	.
\end{equation*}
The upper bound is presented in~\cite{harborth1993upper}.
The lower bound was firstly introduced in~\cite{solymosi2003note};
the largest known value for $c_1$ is $5/11$ for $n\geq 4$~\cite{my-pps-linear-bound-2019}.

@@

In the present paper we give a special bound for planar integral point sets in semi-general position.
The condition of semi-general position is important in the given proof.

@@

\begin{Definition}
	The part of a plane between two parallel straight lines with distance $\rho$ between the lines
	is called a strip of width $\rho$.
\end{Definition}

@@

\begin{lemma}
	\label{lem:count_of_points_on_hyperbolas}
	Let $\{M_1, M_2, M_3, M_4\} \subset M\in\overline{\mathfrak{M}}(2,n)$
	(points $M_2$ and $M_3$ may coincide, other points may not), $n\geq 4$.
	Then $\# M \leq 4 \cdot |M_1 M_2| \cdot |M_3 M_4|$.
\end{lemma}

@@

\begin{proof}
	Each point $N\in M$ satisfies one of the following conditions:

@@

	Due to Lemma~\ref{lem:no_distance_one} we have $|M_1 M_2| \geq 2$ and $|M_3 M_4| \geq 2$.
	Since
	\begin{multline}
		4 (|M_1 M_2| - 1)(|M_3 M_4| - 1) + 4 + 4
		=
		4 ( (|M_1 M_2| - 1)(|M_3 M_4| - 1) + 2)
		=
		\\=
		4 ( |M_1 M_2| \cdot |M_3 M_4| + 1 - |M_1 M_2| - |M_3 M_4| + 2)
		=
		\\=
		4 ( |M_1 M_2| \cdot |M_3 M_4| + (1 - |M_1 M_2|) + (2 - |M_3 M_4|))
		\leq
		4 |M_1 M_2| \cdot |M_3 M_4|
		,
	\end{multline}
	we are done.
\end{proof}

@@

\begin{proof}
	For $n = 3$ we have $\overline{d}(2,3) = 1$.
	Consider $M\in\overline{\mathfrak{M}}(2,n)$, $n \geq 4$, $\operatorname{diam} M = p$.

	Let us choose points $M_1, M_2, M_3, M_4 \in M$
	(points $M_2$ and $M_3$ may coincide, other points may not), such that
	\begin{equation}
		\min_{A, B \in M} |AB| = |M_1 M_2|
		,
	\end{equation}
	\begin{equation}
		\min_{A, B \in M \setminus \{M_1\}} |AB| = |M_3 M_4| = m
		.
	\end{equation}

	For $m \leq p^{2/5}$, Lemma~\ref{lem:count_of_points_on_hyperbolas} yields that
	\begin{equation}
		n \leq 4 \cdot |M_1 M_2| \cdot |M_3 M_4| \leq  4 p^{4/5}
		.
	\end{equation}
	This means that
	\begin{equation}
		\label{eq:hyperbolas_5_4}
		p \geq (n/4) ^ {5/4} > (n/5) ^ {5/4}
		.
	\end{equation}

@@

\begin{Remark}
	The following result in known:
\end{Remark}

\begin{lemma}
	\cite[Corollary 1]{solymosi2003note}
	For $H \in \overline {\mathfrak{M}}(2,n)$, the minimum
	distance in H is at least $n^{1/3}$.
\end{lemma}
Applying the same technique, one can easily derive that
\begin{equation}
	n \leq 3 \frac{\operatorname{diam} H }{n^{1/6}}
	,
\end{equation}
which leads to the bound
\begin{equation}
	\overline{d}(2,n) \geq c_3 n^{7/6}
	,
\end{equation}
which is less than the one from Theorem~\ref{thm:main_result}.

\section{Conclusion}
The presented bound is the first special lower bound for sets in semi-general position.
Thus, we have not accepted the challenge to make the constant in Theorem~\ref{thm:main_result} as large as possible,
in order to keep the ideas of the proof clear and understandable.
A more thorough research can be done in the future to enlarge the constant.
However, the upper and lower bounds are still not tight.

\section{Acknowledgements}
Author thanks Dr. Prof. E.M. Semenov for proofreading and valuable advice.

>>>>

\Abstract
{%Аннотация статьи на русском языке 150-250 слов с учетом ключевых слов
Множество точек $M$ на плоскости называется плоским множеством с целочисленными расстояниями,
если все расстояния между точками $M$ суть целые числа,
и при этом $M$ не содержится ни в какой прямой.
Говорят, что плоское множество с целочисленными расстояниями есть множество полуобщего положения,
если никакие три его точки не лежат на одной прямой.
Известная оценка снизу на диаметр плоского множества с целочисленными расстояниями
линейна относительно его мощности.
Ранее не были известны отдельные оценки снизу на диаметр плоских множеств с целочисленными расстояниями полуобщего положения
заданной мощности
(известная конструктивная оценка сверху на диаметр плоских множеств с целочисленными расстояниями
использует как раз множества полуобщего положения).
В статье доказывается надлинейная оценка снизу
на диаметр плоского множества с целочисленными расстояниями полуобщего положения
(полиномиальная с показателем $5/4$).
Доказательство основано на относительно большом количестве лемм и наблюдений,
включая результаты Солимоси
из статьи, в которой была впервые доказана линейная оценка снизу
на диаметр плоских множеств с целочисленными расстояниями.
Полученная в данной статье оценка всё ещё не точна.
}
{%Аннотация статьи на английском языке
A point set $M$ in the Euclidean plane is said to be a planar integral point set if all the distances between the
elements of $M$ are integers, and $M$ is not situated on a straight line.
A planar integral point set is said to be a set in semi-general position, if it does not contain collinear triples.
The existing lower bound for mininal diameter of planar integral point sets is linear with respect to its cardinality.
There were no known special diameter bounds for planar integral point sets in semi-general position of given cardinality
(the known upper bound for planar integral point sets is constructive
and employs planar integral point sets in semi-general position).
We prove a new lower bound for minimal diameter of planar integral point sets in semi-general position
that is better than linear (polynomial of power $5/4$).
The proof is based on several lemmas and observations, including the ones established by Solymosi
to prove the first linear lower bound for diameter of a planar integral point set.
We have to admit that our bound is still not exact.
}

@@

Since every integral point set can obviously be dilated to a set of larger diameter,
minimal possible diameters of sets of given cardinality are of interest.
To be precise,
the following function was introduced in~\cite{kurz2008bounds,kurz2008minimum}:
\begin{equation}
	d(2,n) = \min_{M\in\mathfrak{M}(2,n)} \operatorname{diam} M
	.
\end{equation}

It turned out to be very easy to construct a planar integral point set of $n$ points with $n-1$ collinear ones and one point out of the line
(so-called \textit{facher} sets);
the same holds for 2 points out of the line (we refer the reader to~\cite{antonov2008maximal}, where some of such sets are called \textit{crabs})
and even for 4 points out of the line~\cite{huff1948diophantine}.
For $9\leq n\leq 122$, the minimal possible diameter is attained at a facher set~\cite{kurz2008bounds}.

@@

Furthermore, the constructions of integral point sets in semi-general position of arbitrary cardinality
appeared in~\cite{harborth1993upper};
such sets are situated on a circle.
Also, there is a sophisticated construction of a circular integral point set of arbitrary cardinality
that gives the possible numbers of odd integral distances
between points in the plane~\cite{piepmeyer1996maximum}.

\begin{Definition}
	A set $M\in\overline{\mathfrak{M}}(2,n)$ is said to be a set in \textit{general position},
	if no four points of $M$ are concyclic.
	The set of all planar integral point sets in general position
	is denoted by $\dot{\mathfrak{M}}(2,n)$.
\end{Definition}

It remains unknown whether there are integral points sets in general position of arbitrary cardinality;
however, some sets $M\in \dot{\mathfrak{M}}(2,7)$ are known~\cite{kreisel2008heptagon,kurz2013constructing}.

The inequality
\begin{equation*}
	d(2,n) \leq \overline{d}(2,n) \leq \dot{d}(2,n)
	,
\end{equation*}
where
$
	\overline{d}(2,n) = \min_{M\in\overline{\mathfrak{M}}(2,n)} \operatorname{diam} M
$
and
$
	\dot{d}(2,n) = \min_{M\in\dot{\mathfrak{M}}(2,n)} \operatorname{diam} M
$,
is obvious; however, a more interesting relation holds:
\begin{equation*}
	c_1 n \leq d(2,n) \leq \overline{d}(2,n) \leq n^{c_2 \log \log n}
	.
\end{equation*}
The upper bound is established in~\cite{harborth1993upper}.
The lower bound firstly appeared in~\cite{solymosi2003note};
the largest known value for $c_1$ is $5/11$ for $n\geq 4$~\cite{my-pps-linear-bound-2019}.

@@

In the present paper we give a special bound for planar integral point sets in semi-general position.
The condition of semi-general position is crucial in the given proof.

@@

\begin{Definition}
	The part of a plane between two parallel straight lines with distance $\rho$ between them
	is called a strip of width $\rho$.
\end{Definition}

@@

\begin{lemma}
	\label{lem:count_of_points_on_hyperbolas}
	Let $\{M_1, M_2, M_3, M_4\} \subset M\in\overline{\mathfrak{M}}(2,n)$
	(points $M_2$ and $M_3$ may coincide, other points are distinct), $n\geq 4$.
	Then $\# M \leq 4 \cdot |M_1 M_2| \cdot |M_3 M_4|$.
\end{lemma}

@@

\begin{proof}[Proof]
	Each point $N\in M$ satisfies one of the following conditions:

@@

	Due to Lemma~\ref{lem:no_distance_one} we have $|M_1 M_2| \geq 2$ and $|M_3 M_4| \geq 2$.
	Since
	\begin{multline}
		4 (|M_1 M_2| - 1)(|M_3 M_4| - 1) + 4 + 4
		=
		4 ( (|M_1 M_2| - 1)(|M_3 M_4| - 1) + 2)
		=
		\\=
		4 ( |M_1 M_2| \cdot |M_3 M_4| + 1 - |M_1 M_2| - |M_3 M_4| + 2)
		=
		\\=
		4 ( |M_1 M_2| \cdot |M_3 M_4| + (1 - |M_1 M_2|) + (2 - |M_3 M_4|))
		\leq
		4 |M_1 M_2| \cdot |M_3 M_4|
		,
	\end{multline}
	the assertion follows.
\end{proof}

@@

\begin{proof}[Proof]
	For $n = 3$ we have $\overline{d}(2,3) = 1$.
	Consider $M\in\overline{\mathfrak{M}}(2,n)$, $n \geq 4$, $\operatorname{diam} M = p$.

	Let us choose points $M_1, M_2, M_3, M_4 \in M$
	(points $M_2$ and $M_3$ may coincide, other points are distinct), such that
	\begin{equation}
		\min_{A, B \in M} |AB| = |M_1 M_2|
		,
	\end{equation}
	\begin{equation}
		\min_{A, B \in M \setminus \{M_1\}} |AB| = |M_3 M_4| = m
		.
	\end{equation}

	For $m \leq p^{2/5}$, Lemma~\ref{lem:count_of_points_on_hyperbolas} yields that
	\begin{equation}
		n \leq 4 \cdot |M_1 M_2| \cdot |M_3 M_4| \leq  4 p^{4/5}
		.
	\end{equation}
	Equivalently,
	\begin{equation}
		\label{eq:hyperbolas_5_4}
		p \geq (n/4) ^ {5/4} > (n/5) ^ {5/4}
		.
	\end{equation}

@@

\begin{Remark}
	The following result appeared in~\cite{solymosi2003note} as Corollary 1:
\end{Remark}

\begin{lemma}
	For $H \in \overline {\mathfrak{M}}(2,n)$, the minimal
	distance in H is at least $n^{1/3}$.
\end{lemma}
Applying the same technique, one can easily derive that
\begin{equation}
	n \leq 3 \frac{\operatorname{diam} H }{n^{1/6}}
	,
\end{equation}
which leads to the bound
\begin{equation}
	\overline{d}(2,n) \geq c_3 n^{7/6}
	,
\end{equation}
which is less than the one from Theorem~\ref{thm:main_result}.

\section{Conclusion}
The presented bound is the first special lower bound for sets in semi-general position.
Thus, we have not accepted the challenge to make the constant in Theorem~\ref{thm:main_result} as large as possible,
in order to keep the ideas of the proof clear and understandable.
The more thorough research can be done in the future to improve the constant.
However, the upper and lower bounds are still not exact.

\section{Acknowledgements}
Author thanks Dr. Prof. E.M. Semenov for proofreading and valuable advices and Dr. A.S. Usachev for proofreading.
